Appendix to 'cubble: An R Package for Organizing and Wrangling Multivariate Spatio-temporal Data

### 2022-07-09 1-intro.Rmd
check on \proglang, \pkg, \code being used throughout the paper

The SpatioTemporal CRAN task view [@ctvspatiotemporal] gathers information about R packages designed for spatio-temporal data and it has a section on *Representing data* that lists existing spatio-temporal data representations used in \proglang{R}. Among them, @spacetime summarises spatio-temporal data into three forms: time-wide, space-wide, and long formats. The associated package \pkg{spacetime} [@spacetime] implements four spatio-temporal layouts (full grid, sparse grid, irregular, and trajectory) to handle different space and time combinations. The \pkg{stars} [@stars] package has a new implementation to use dense arrays to represent spatio-temporal cubes. It also interfaces with \pkg{sf} [@sf], a package commonly used for wrangling spatial data, and the \pkg{tidyverse} [@tidyverse] suite for general data wrangling and visualization in \proglang{R}.

@@

This paper presents a new \proglang{R} package, \pkg{cubble} which addresses the two issues mentioned above. In the package, a new data structure, also called \code{cubble}, is proposed to organize spatial and temporal variables as two forms of a single data object so that they can be wrangled separately or combined while being kept synchronized. Among the four spacetime layouts in @spacetime, \pkg{cubble} can be applied to full grid, sparse grid, or irregular, but not trajectory, which is outside the scope of this work. The software is available from the Comprehensive R Archive Network (CRAN) at https://CRAN.R-project.org/package=cubble.

>>>>

The SpatioTemporal CRAN task view [@ctvspatiotemporal] gathers information about \proglang{R} packages designed for spatio-temporal data and it has a section on *Representing data* that lists existing spatio-temporal data representations used in \proglang{R}. Among them, @spacetime summarises spatio-temporal data into three forms: time-wide, space-wide, and long formats. The associated package \pkg{spacetime} [@spacetime] implements four spatio-temporal layouts (full grid, sparse grid, irregular, and trajectory) to handle different space and time combinations. The \pkg{stars} [@stars] package has a new implementation to use dense arrays to represent spatio-temporal cubes. It also interfaces with \pkg{sf} [@sf], a package commonly used for wrangling spatial data, and the \pkg{tidyverse} [@tidyverse] suite for general data wrangling and visualization in \proglang{R}.

@@

This paper presents a new \proglang{R} package, \pkg{cubble} which addresses the two issues mentioned above. In the package, a new data structure, also called \code{cubble}, is proposed to organize spatial and temporal variables as two forms of a single data object so that they can be wrangled separately or combined while being kept synchronized. Among the four spacetime layouts in @spacetime, \pkg{cubble} can be applied to full grid, sparse grid, or irregular, but not trajectory, which is outside the scope of this work. The software is available from the Comprehensive \proglang{R} Archive Network (CRAN) at https://CRAN.R-project.org/package=cubble.

### 2022-07-09 3-the-cubble-pkg.Rmd
check on \proglang, \pkg, \code being used throughout the paper

A cubble can be created from various common spatio-temporal data formats, including basic R objects like tibble, tsibble and sf. Section \@ref(covid) describes converting multiple tables into a cubble object and Section \@ref(netcdf) illustrates how to convert a netCDF object.

@@

NetCDF data is another format commonly used for storing spatio-temporal data. It has two main components: *dimension* for defining the spatio-temporal grid (longitude, latitude, and time) and *variable* that populates the defined grid. Attributes  can be associated with dimensions or variables. Because there can be many different styles of representing this information there is a metadata convention [@hassell_data_2017] to standardize the format of the attributes. A few packages in R exist for manipulating NetCDF data and these include a high-level R interface: \pkg{ncdf4} [@ncdf4], a low-level interface that calls a C-interface: \pkg{RNetCDF} [@rnetcdf], and a tidyverse implementation: \pkg{tidync} [@tidync].

>>>>

A cubble can be created from various common spatio-temporal data formats, including basic \proglang{R} objects like tibble, tsibble and sf. Section \@ref(covid) describes converting multiple tables into a cubble object and Section \@ref(netcdf) illustrates how to convert a netCDF object.

@@

NetCDF data is another format commonly used for storing spatio-temporal data. It has two main components: *dimension* for defining the spatio-temporal grid (longitude, latitude, and time) and *variable* that populates the defined grid. Attributes  can be associated with dimensions or variables. Because there can be many different styles of representing this information there is a metadata convention [@hassell_data_2017] to standardize the format of the attributes. A few packages in \proglang{R} exist for manipulating NetCDF data and these include a high-level R interface: \pkg{ncdf4} [@ncdf4], a low-level interface that calls a C-interface: \pkg{RNetCDF} [@rnetcdf], and a tidyverse implementation: \pkg{tidync} [@tidync].

### 2022-07-09 4-other-features.Rmd
check on \proglang, \pkg, \code being used throughout the paper

The cubble workflow fits works well with an interactive graphics pipeline (e.g., @buja1988elements, @buja1996interactive,  @sutherland2000orca, @xie2014reactive, @cheng2016enabling) that is available in R with the package crosstalk [@crosstalk]. Figure \ref{fig:illu-interactive} illustrates how linking can be achieved between a map and multiple time series using a cubble. The map (produced from the nested form) and time series (produced from the long form) are both shared crosstalk objects. When a user makes a selection on the map, the site is highlighted (a). This activates a row in the nested cubble object, which is then communicated to the long cubble -- all the observations with the same \code{id} (b) will be selected. The long cubble will then highlight the corresponding series in the time series plot (c).

>>>>

The cubble workflow fits works well with an interactive graphics pipeline (e.g., @buja1988elements, @buja1996interactive,  @sutherland2000orca, @xie2014reactive, @cheng2016enabling) that is available in \proglang{R} with the package crosstalk [@crosstalk]. Figure \ref{fig:illu-interactive} illustrates how linking can be achieved between a map and multiple time series using a cubble. The map (produced from the nested form) and time series (produced from the long form) are both shared crosstalk objects. When a user makes a selection on the map, the site is highlighted (a). This activates a row in the nested cubble object, which is then communicated to the long cubble -- all the observations with the same \code{id} (b) will be selected. The long cubble will then highlight the corresponding series in the time series plot (c).

### 2022-07-09 5-2-aus-hist-max-temp.Rmd
check on \proglang, \pkg, \code being used throughout the paper

The Global Historical Climatology Network (GHCN) provides daily climate measures from stations across the world. The data used here (\code{historical_tmax}) is a subset extracted using the `rnoaa` package, containing the records of maximum temperature for `r nrow(historical_tmax)` Australian stations from `r range(historical_tmax$first_year)[1]` through `r range(historical_tmax$first_year)[2]` and provides information also on the latitude, longitude and elevation of each of the  stations. <!-- This data  is already cast into a \code{cubble} data object (please see Appendix X for the code to extract  the data from XX and prepare the data into \code{cubble} form), with \code{id} as the key, \code{date} as the index, and \code{c(longitude, latitude)} as the coordinates.--> The goal of this example is to compare the monthly average maximum temperature between two periods, 1971-1975 and 2016-2020, for stations in Victoria and New South Wales, using a glyph map.

>>>>

The Global Historical Climatology Network (GHCN) provides daily climate measures from stations across the world. The data used here (\code{historical_tmax}) is a subset extracted using the \pkg{rnoaa}[@rnoaa] package, containing the records of maximum temperature for `r nrow(historical_tmax)` Australian stations from `r range(historical_tmax$first_year)[1]` through `r range(historical_tmax$first_year)[2]` and provides information also on the latitude, longitude and elevation of each of the  stations. <!-- This data  is already cast into a \code{cubble} data object (please see Appendix X for the code to extract  the data from XX and prepare the data into \code{cubble} form), with \code{id} as the key, \code{date} as the index, and \code{c(longitude, latitude)} as the coordinates.--> The goal of this example is to compare the monthly average maximum temperature between two periods, 1971-1975 and 2016-2020, for stations in Victoria and New South Wales, using a glyph map.

### 2022-07-09 5-4-era5.Rmd
check on \proglang, \pkg, \code being used throughout the paper

The ERA5 data [@hersbach2020era5] provides hourly estimates across the Earth for atmospheric, land and oceanic climate variables. The data is available in the NetCDF format from the European Centre for Medium-Range Weather Forecasts (ECMWF). It can be directly downloaded from Copernicus Climate Data Store (CDS) [@cds] website or via the R package \pkg{ecmwfr} [@ecwmfr].

@@

<!--The goal of this example is to reproduce the break-up of the southern polar vortex that happened in late September and early October in 2002 in the stratosphere in the Southern Hemisphere. Therefore, we filter data corresponding to the Southern Hemisphere and restrict our analysis to four dates: 2002-09-22, 2002-09-26, 2002-09-30, and 2002-10-04.--> The downloaded netCDF data is read into R using the \code{ncdf4} package, and converted to a \code{cubble}:

>>>>

The ERA5 data [@hersbach2020era5] provides hourly estimates across the Earth for atmospheric, land and oceanic climate variables. The data is available in the NetCDF format from the European Centre for Medium-Range Weather Forecasts (ECMWF). It can be directly downloaded from Copernicus Climate Data Store (CDS) [@cds] website or via the \proglang{R} package \pkg{ecmwfr} [@ecwmfr].

@@

<!--The goal of this example is to reproduce the break-up of the southern polar vortex that happened in late September and early October in 2002 in the stratosphere in the Southern Hemisphere. Therefore, we filter data corresponding to the Southern Hemisphere and restrict our analysis to four dates: 2002-09-22, 2002-09-26, 2002-09-30, and 2002-10-04.--> The downloaded netCDF data is read into \proglang{R} using the \code{ncdf4} package, and converted to a \code{cubble}:

### 2022-07-09 6-conclusion.Rmd
check on \proglang, \pkg, \code being used throughout the paper

This paper presents an \proglang{R} package \pkg{cubble} for organizing, manipulating and visualizing spatio-temporal data. The package introduces a new data structure for spatio-temporal data, \code{cubble}, that connects the time invariant and varying variables and that allows the user to work with a nested and long form of the data. This work adds capabilities into the spatio-temporal practitioners toolbox to integrate it with a tidy data framework. The data structure and functions introduced in this package can be used and combined with existing spatio-temporal R packages such as \pkg{sf}, data wrangling packages such as \pkg{dplyr}, and visualization packages such as \pkg{ggplot2}, \pkg{plotly} and \pkg{leaflet}.

@@

This work is funded by a Commonwealth Scientific and Industrial Research Organisation (CSIRO) Data61 Scholarship and started while Nicolas Langrené was affiliated with CSIRO's Data61. The article is created using \pkg{knitr} [@knitr] and \pkg{rmarkdown} [@rmarkdown] in R using the `rticles::jss_article` template. The source code for reproducing this paper can be found at: https://github.com/huizezhang-sherry/paper-cubble.

>>>>

This paper presents an \proglang{R} package \pkg{cubble} for organizing, manipulating and visualizing spatio-temporal data. The package introduces a new data structure for spatio-temporal data, \code{cubble}, that connects the time invariant and varying variables and that allows the user to work with a nested and long form of the data. This work adds capabilities into the spatio-temporal practitioners toolbox to integrate it with a tidy data framework. The data structure and functions introduced in this package can be used and combined with existing spatio-temporal \proglang{R} packages such as \pkg{sf}, data wrangling packages such as \pkg{dplyr}, and visualization packages such as \pkg{ggplot2}, \pkg{plotly} and \pkg{leaflet}.

@@

This work is funded by a Commonwealth Scientific and Industrial Research Organisation (CSIRO) Data61 Scholarship and started while Nicolas Langrené was affiliated with CSIRO's Data61. The article is created using \pkg{knitr} [@knitr] and \pkg{rmarkdown} [@rmarkdown] in R using the \code{rticles::jss_article} template. The source code for reproducing this paper can be found at: https://github.com/huizezhang-sherry/paper-cubble.

### 2022-07-09 jss/appendix.Rmd
check on \proglang, \pkg, \code being used throughout the paper

In  example 5.2, some overlapping of the glyphs occurred for a number of stations on the right hand side of the map in Figure 6. This is a problem when mapping time series or other glyphs corresponding to locations that are geographically closed on the map. In some cases it is better to display such information at an aggregated level by grouping the data adequately before exhibiting the information on a figure. The example below shows how can  spatio-temporal data be organised at different hierarchical levels so that information can be grouped both temporal and spatially using  `switch_key()`. The goal of this example is to highlight how easy is to move across the data hierarchy using **cubble**.

>>>>

In  example 5.2, some overlapping of the glyphs occurred for a number of stations on the right hand side of the map in Figure 6. This is a problem when mapping time series or other glyphs corresponding to locations that are geographically closed on the map. In some cases it is better to display such information at an aggregated level by grouping the data adequately before exhibiting the information on a figure. The example below shows how can  spatio-temporal data be organised at different hierarchical levels so that information can be grouped both temporal and spatially using  \code{switch_key()}. The goal of this example is to highlight how easy is to move across the data hierarchy using **cubble**.

### 2022-07-09 jss/main.Rmd
check on \proglang, \pkg, \code being used throughout the paper

---
documentclass: jss
author:
  - name: H. Sherry Zhang
    affiliation: Monash University
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email: \email{huize.zhang@monash.edu}
  - name: Dianne Cook
    affiliation: 'Monash University \AND'
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email: dicook@monash.edu
  - name: Ursula Laa
    affiliation: 'University of Natural Resources and Life Sciences \AND'
    address: Gregor-Mendel-Straße 33, 1180 Wien, Austria
    email:  ursula.laa@boku.ac.at
  - name: Nicolas Langrené
    affiliation: 'BNU-HKBU United International College'
    address: 2000 Jintong Road, Tangjiawan, Zhuhai, Guangdong Province, China
    email: nicolaslangrene@uic.edu.cn
  - name: Patricia Menéndez
    affiliation: 'Monash University'
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email:  patricia.menendez@monash.edu
title:
  formatted: "\\pkg{cubble}: An R Package for Organizing and Wrangling Multivariate Spatio-temporal Data"
  # If you use tex in the formatted title, alsoge supply version without
  plain:     "cubble: An R Package for Organizing and Wrangling Multivariate Spatio-temporal Data"
  # For running headers, if needed
  # short:     "\\pkg{foo}: A Capitalized Title"
abstract: >
     Multivariate spatio-temporal data refers to multiple measurements taken across space and time. For many analyses, spatial and time components can be separately studied: for example, to explore the temporal trend of one variable for a single spatial location, or to model the spatial distribution of one variable at a given time. However for some studies, it is important to analyze different aspects of the spatio-temporal data simultaneously, like for instance, temporal trends of multiple variables across locations. In order to facilitate the study of different portions or combinations of spatio-temporal data, we introduce a new data structure, \code{cubble}, with a suite of functions enabling easy slicing and dicing on the different components spatio-temporal components.  The proposed \code{cubble} structure ensures that all the components of the data are easy to access and manipulate while providing flexibility for data analysis. In addition, \code{cubble} facilitates visual and numerical explorations of the data while easing data wrangling and modelling. The \code{cubble} structure and the functions provided in the \code{cubble} R package equip users with the capability to handle hierarchical spatial and temporal structures. The \code{cubble} structure and the tools implemented in the package are illustrated with different examples of Australian climate data.
keywords:
  # at least one keyword must be supplied
  formatted: [spatial, temporal, spatio temporal, "\\proglang{R}", environmental data, exploratory data analysis]
  plain:     [spatial, temporal, spatio temporal, R, environmental data, exploratory data analysis]
preamble: >
  \usepackage{amsmath}
  \usepackage{array}
  \usepackage{xcolor}
  \newcommand{\sherry}[1]{\textcolor{magenta}{#1}}
output: 
  bookdown::pdf_book:
    base_format: rticles::jss_article
bibliography: ../references.bib
---

>>>>

---
documentclass: jss
author:
  - name: H. Sherry Zhang
    affiliation: Monash University
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email: \email{huize.zhang@monash.edu}
  - name: Dianne Cook
    affiliation: 'Monash University \AND'
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email: dicook@monash.edu
  - name: Ursula Laa
    affiliation: 'University of Natural Resources and Life Sciences \AND'
    address: Gregor-Mendel-Straße 33, 1180 Wien, Austria
    email:  ursula.laa@boku.ac.at
  - name: Nicolas Langrené
    affiliation: 'BNU-HKBU United International College'
    address: 2000 Jintong Road, Tangjiawan, Zhuhai, Guangdong Province, China
    email: nicolaslangrene@uic.edu.cn
  - name: Patricia Menéndez
    affiliation: 'Monash University'
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email:  patricia.menendez@monash.edu
title:
  formatted: "\\pkg{cubble}: An \proglang{R} Package for Organizing and Wrangling Multivariate Spatio-temporal Data"
  # If you use tex in the formatted title, alsoge supply version without
  plain:     "cubble: An R Package for Organizing and Wrangling Multivariate Spatio-temporal Data"
  # For running headers, if needed
  # short:     "\\pkg{foo}: A Capitalized Title"
abstract: >
     Multivariate spatio-temporal data refers to multiple measurements taken across space and time. For many analyses, spatial and time components can be separately studied: for example, to explore the temporal trend of one variable for a single spatial location, or to model the spatial distribution of one variable at a given time. However for some studies, it is important to analyze different aspects of the spatio-temporal data simultaneously, like for instance, temporal trends of multiple variables across locations. In order to facilitate the study of different portions or combinations of spatio-temporal data, we introduce a new data structure, \code{cubble}, with a suite of functions enabling easy slicing and dicing on the different components spatio-temporal components.  The proposed \code{cubble} structure ensures that all the components of the data are easy to access and manipulate while providing flexibility for data analysis. In addition, \code{cubble} facilitates visual and numerical explorations of the data while easing data wrangling and modelling. The \code{cubble} structure and the functions provided in the \code{cubble} \proglang{R} package equip users with the capability to handle hierarchical spatial and temporal structures. The \code{cubble} structure and the tools implemented in the package are illustrated with different examples of Australian climate data.
keywords:
  # at least one keyword must be supplied
  formatted: [spatial, temporal, spatio temporal, "\\proglang{R}", environmental data, exploratory data analysis]
  plain:     [spatial, temporal, spatio temporal, R, environmental data, exploratory data analysis]
preamble: >
  \usepackage{amsmath}
  \usepackage{array}
  \usepackage{xcolor}
  \newcommand{\sherry}[1]{\textcolor{magenta}{#1}}
output: 
  bookdown::pdf_book:
    base_format: rticles::jss_article
bibliography: ../references.bib
---
